---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# LogRatioReg: Log-ratio Lasso Regression

<!-- badges: start -->
[![R-CMD-check](https://github.com/tengfei-emory/LogRatioReg/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tengfei-emory/LogRatioReg/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`LogRatioReg` is an open-source computational tool to perform log-ratio lasso regression modeling and microbial feature selection for continuous, binary, time-to-event, and competing risk outcomes. The proposed method adapts the augmented Lagrangian algorithm for a zero-sum constraint optimization problem while enabling a two-stage screening process for extended false-positive control.

## Installation

You can install the development version of `LogRatioReg` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tengfei-emory/LogRatioReg")
```

## Example

Here is a toy example for linear regression with 10-fold cross-validation with `LogRatioLasso()` for a simulated data with 50 samples and 100 compositional features.

The data simulation procedure is described in the preprint.

```{r example}

library(LogRatioReg)

dat <- simu(n=50,p=100,model="linear")
fit <- LogRatioLasso(dat$x,dat$y,ncv=10,progress=FALSE)

```

To view plots of cross-validated prediction error and parameter coefficients, use `fit$pmse` or `fit$pcoef`:

```{r plot, echo = FALSE,fig.height=3,fig.width=6}
fit$pmse
fit$pcoef
```

For binary and survival outcomes, please use the corresponding functions `LogRatioLogisticLasso()` and `LogRatioCoxLasso()`. 

```{r example2}

library(LogRatioReg)
library(survival)

dat.bin <- simu(n=50,p=100,model="binomial")
fit.bin <- LogRatioLogisticLasso(dat.bin$x,dat.bin$y,ncv=10,progress=FALSE)

dat.cox <- simu(n=50,p=100,model="cox")
fit.cox <- LogRatioCoxLasso(dat.cox$x,Surv(dat.cox$t,dat.cox$d),ncv=10,progress=FALSE)

```