---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# FLORAL: Fit LOg-RAtio Lasso regression for compositional covariates

<!-- badges: start -->
[![R-CMD-check](https://github.com/vdblab/FLORAL/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/vdblab/FLORAL/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/vdblab/FLORAL/branch/master/graph/badge.svg)](https://app.codecov.io/gh/vdblab/FLORAL?branch=master)
<!-- badges: end -->

`FLORAL` is an open-source computational tool to perform log-ratio lasso regression modeling and microbial feature selection for continuous, binary, time-to-event, and competing risk outcomes. The proposed method adapts the augmented Lagrangian algorithm for a zero-sum constraint optimization problem while enabling a two-stage screening process for extended false-positive control.

## Installation

You can install the development version of `FLORAL` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("vdblab/FLORAL")
```

## Example

Here is a toy example for linear regression with 10-fold cross-validation for a simulated data with 50 samples and 100 compositional features.

The data simulation procedure is described in the preprint.

```{r example}
set.seed(23420)
library(FLORAL)

dat <- simu(n=50,p=100,model="linear")
fit <- FLORAL(dat$xcount,dat$y,family="gaussian",ncv=10,progress=FALSE)
```

To view plots of cross-validated prediction error and parameter coefficients, use `fit$pmse` or `fit$pcoef`:

```{r plot, echo = FALSE,fig.height=3,fig.width=6,dpi=300}
fit$pmse
fit$pcoef
```

To view selected compositional features, use `fit$selected.feature`, where features are sorted by their names. Features under `min` and `1se` correspond to penalty parameter $\lambda_{\min}$ and $\lambda_{\text{1se}}$, respectively. Features under `min.2stage` and `1se.2stage` are obtained after applying 2-stage filtering based on features under `min` and `1se`, respectively.

```{r feature}
fit$selected.feature
```

To get specific ratios selected by the 2-stage procedure, use `fit$step2.ratios`, where `min` and `1se` display the ratios between features. For each identified ratio, `min.idx` and `1se.idx` return the column indices in the original input matrix for the two corresponding features forming the ratio.

```{r ratio}
fit$step2.ratios
```

For binary and survival outcomes, please specify `family="binomial"` and `family="cox"` accordingly. 

```{r example2, warning=FALSE}
dat.bin <- simu(n=50,p=100,model="binomial")
fit.bin <- FLORAL(dat.bin$xcount,dat.bin$y,family="binomial",ncv=10,progress=FALSE)

dat.cox <- simu(n=50,p=100,model="cox")
fit.cox <- FLORAL(dat.cox$xcount,survival::Surv(dat.cox$t,dat.cox$d),family="cox",ncv=10,progress=FALSE)
```
