---
title: "Using FLORAL with phyloseq data"
output:
  rmarkdown::html_vignette:
    md_extensions: [ 
      "-autolink_bare_uris" 
    ]
vignette: >
  %\VignetteIndexEntry{Using FLORAL with phyloseq data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(phyloseq)
```

[Phyloseq](http://joey711.github.io/phyloseq/) is a popular package for working with microbiome data.  Here we show how to use the `phy_to_floral_data` helper function to convert phyloseq data into a format accepted by FLORAL.

The following code downloads data described in [this paper](https://www.nature.com/articles/s41597-021-00860-8) and turns it into a phyloseq object.  The tax_glom step here takes some time, and can be replaced with [`speedyseq::tax_glom`](https://github.com/mikemc/speedyseq) for better performace.
```{r}
#this file has duplicate rows, and has multiple rows per pool
samples <- read.csv("https://figshare.com/ndownloader/files/33076496") %>% distinct() %>% 
  select(-Pool, -Run, -ShotgunBatchID) %>% distinct()
samples <- samples[1:100,] # Using the first 100 samples only.
counts <- read.csv("https://figshare.com/ndownloader/files/26393788") 
counts <- counts %>%
  filter(SampleID %in% samples$SampleID)
taxonomy <- read.csv("https://figshare.com/ndownloader/files/26770997")

phy <- phyloseq(
  sample_data(samples %>% column_to_rownames("SampleID")),
  tax_table(taxonomy %>% select(ASV, Kingdom:Genus) %>% column_to_rownames("ASV") %>% as.matrix()),
  otu_table(counts  %>% pivot_wider(names_from = "SampleID", values_from = "Count", values_fill = 0) %>% column_to_rownames("ASV") %>% as.matrix(), taxa_are_rows = TRUE)
) %>%  subset_samples(DayRelativeToNearestHCT > -30 & DayRelativeToNearestHCT < 0) %>% 
  tax_glom("Genus")
```

Next, we convert that phyloseq object into a list of results to be used by FLORAL; we have to specify the main outcome of interest as `y`, and any metadata columns (from `sample_data(phy)`) to use as covariates.  Note that the analysis described here is just an example for using the function; this


```{r}
dat <- FLORAL::phy_to_floral_data(phy, covariates=c("Consistency"), y = "DayRelativeToNearestHCT")
```

The resulting list has named entities for the main arguments to FLORAL:

```{r}
res <- FLORAL::FLORAL(y = dat$y, x = dat$xcount, ncov = dat$ncov, family = "gaussian", ncv=NULL)
```